---
title: "Untitled"
author: "Yara Ghabra 1006336056, Grace Walker Mitchell 1005046326"
date: "2022-11-22"
output: html_document
---


```{r}
library(tidyverse)
library(lmerTest)
library(lme4)
library(ggalt)
library(MuMIn)
library(sjmisc)
library(car)
```

```{r}
#writing datasets into CSV for sharing in github
write.csv(pheno_data, "final_project_full_data.csv")
write.csv(spp_filt, "final_project_filtered_data.csv")
```


###Data Manipulation
```{r}
#read in the unaltered phenophase data set
pheno_data<-read.csv("/Users/gracewalkermitchell/Desktop/EEB313/project/CCIN13215_20220120_tundra_phenology_database.csv")

#filter the dataset by flowering phenophase, control treatment, forb functional group, and species of interest (species that occur over the same time span with a lot of observations)
pheno_data_filt<-pheno_data %>% 
  filter(phenophase == "flower") %>% 
  filter(treatment == "CTL") %>% 
  filter(functional_group == "forb") %>% 
  filter(spp == "CARBEL" | spp == "DRALAC" | spp == "PAPRAD" | spp == "POLVIV" | spp == "PEDHIR" | spp == "SAXOPP" | spp == "SILACA" | spp == "STECRA")

#seeing how many species occur in both wet and dry
numbers<-pheno_data_filt %>% 
  group_by(spp,soil_moisture) %>% 
  summarise(n=n()) %>% 
  ungroup() %>% 
  filter(soil_moisture == "wet" | soil_moisture =="dry" | soil_moisture == "moist")
  
#filtering by four species of interest - occur in both wet and dry, have a similar time range of observations, and have a lot of observations
spp_filt<-pheno_data_filt %>% 
  filter(spp == "DRALAC" | spp == "PAPRAD" | spp == "POLVIV" | spp == "PEDHIR") %>% 
  filter(soil_moisture =="wet" | soil_moisture =="moist" | soil_moisture =="dry") %>% 
  filter(study_area != "Val Bercla") #excluding this because it is an alpine site, not an arctic site

#only including wet and dry, not moist
spp_filt[spp_filt == "moist"] <- "wet"
```

### T tests

```{r}
#creating a separate dataset for each species for t tests
polviv_ttest<-spp_filt %>% 
  filter(spp == "POLVIV")
dralac_ttest<-spp_filt %>% 
  filter(spp == "DRALAC")
paprad_ttest<-spp_filt %>% 
  filter(spp == "PAPRAD")
pedhir_ttest<-spp_filt %>% 
  filter(spp == "PEDHIR")

#removing blank values from the POLVIV dataset
polviv_ttest<-polviv_ttest %>% 
  filter(soil_moisture == "dry" | soil_moisture == "wet")
```

T tests have two assumptions:
1. The dependent variable must be normally distributed in both samples
2. The variance of the dependent variable must be approximately equal between the two samples

Shapiro-wilk's test:

in this case if we get a non-significant p-value, our data is normally distributed

```{r  initial test of t test assumptions}
#Shapiro-wilk's test to test normality
shapiro.test(paprad_ttest$DOY)
#not normally distributed
shapiro.test(polviv_ttest$DOY)
#not normally distributed
shapiro.test(pedhir_ttest$DOY)
#not normally distributed
shapiro.test(dralac_ttest$DOY)
#normally distributed

# F-test to test for homogeneity in variances
# H0: Ratio of variance is equal to 1, i.e., samples have equal variance
var.test(DOY ~ soil_moisture, data=paprad_ttest)
var.test(DOY ~ soil_moisture, data=polviv_ttest)
var.test(DOY ~ soil_moisture, data=pedhir_ttest)
var.test(DOY ~ soil_moisture, data=dralac_ttest)
```
```{r  data transformations for normality}
#transforming data to achieve normal distribution
hist(spp_filt$DOY) #right skewed

#performing log base 10 transformation
spp_filt<- spp_filt %>% 
  mutate(inv_DOY= 1/DOY , log10_DOY= log10(DOY) , sqrt_DOY= sqrt(DOY) , log_DOY= log(DOY))

#---------------------------------------------------------------
#have to run this again to update
polviv_ttest<-spp_filt %>% 
  filter(spp == "POLVIV")
dralac_ttest<-spp_filt %>% 
  filter(spp == "DRALAC")
paprad_ttest<-spp_filt %>% 
  filter(spp == "PAPRAD")
pedhir_ttest<-spp_filt %>% 
  filter(spp == "PEDHIR")

#removing blank values from the POLVIV dataset
polviv_ttest<-polviv_ttest %>% 
  filter(soil_moisture == "dry" | soil_moisture == "wet")
#---------------------------------------------------------------

hist(polviv_ttest$inv_DOY) #normal now
hist(dralac_ttest$inv_DOY) #normalish -- doesn't need transformations
hist(pedhir_ttest$inv_DOY) #left skewed
hist(paprad_ttest$log10_DOY) #normal

#checking Shapiro-wilk's test again:
shapiro.test(paprad_ttest$log10_DOY)
#not normally distributed
shapiro.test(polviv_ttest$inv_DOY)
#not normally distributed
shapiro.test(pedhir_ttest$inv_DOY)
#not normally distributed
shapiro.test(dralac_ttest$DOY)
#normally distributed


#discuss that there are errors in our results cuz data is not significatly normal and dont have homogeniety of variances
```

```{r f test after transformations}
#F tests after data transformations
var.test(log10_DOY ~ soil_moisture, data=paprad_ttest)
var.test(inv_DOY ~ soil_moisture, data=polviv_ttest)
var.test(inv_DOY ~ soil_moisture, data=pedhir_ttest)
# var.test(DOY ~ soil_moisture, data=dralac_ttest)
```


```{r}
#performing t test for paprad
t.test(log10_DOY~soil_moisture, data = paprad_ttest)
```

```{r}
#t test for polviv
t.test(inv_DOY~soil_moisture, data = polviv_ttest)
```

```{r}
# t test for dralac
t.test(DOY~soil_moisture, data = dralac_ttest)
```


```{r}
#t test for pedhir
t.test(inv_DOY~soil_moisture, data = pedhir_ttest)
```

### Modelling 

```{r}
#this is just for f-ing around, dont include
t.test(DOY~soil_moisture, data=spp_filt)
summary(aov(DOY~spp, data=spp_filt))

fixed_model<- lm(DOY~soil_moisture*spp, data=spp_filt)
summary(fixed_model)

```

```{r}
#mixed models

mix_model_int1<-lmer(DOY~soil_moisture * spp +
                      (1|subsite),
                    data = spp_filt)

mix_model_int2<-lmer(DOY~soil_moisture * spp +
                      (1|study_area),
                    data = spp_filt)

mix_model_int3<-lmer(DOY~soil_moisture * spp +
                      (1|study_area) + (1|subsite),
                    data = spp_filt)

mix_model_int4<-lmer(DOY~soil_moisture * spp + year +
                      (1|study_area) + (1|subsite),
                    data = spp_filt)

mix_model_int5<- lmer(DOY~soil_moisture * spp * year +
                      (1|subsite),
                    data = spp_filt)

mix_model_int6<-lmer(DOY~soil_moisture * spp * year +
                      (1|study_area),
                    data = spp_filt)
```


```{r}
#creating a fixed model to test the effects of soil moisture on flowering date

fixed_model2<-lm(DOY~soil_moisture * spp + subsite,
                    data = spp_filt)
summary(fixed_model2)

fixed_model3<-lm(DOY~soil_moisture * spp + study_area,
                    data = spp_filt)
summary(fixed_model2)
```

```{r}
#getting DF

Anova(mix_model_int3, p.adjust.methods = TRUE, type = 2) # in package car to get degrees of freedoms
```


```{r}
#Using AICc to compare all models

AICc(mix_model_int1, mix_model_int2, mix_model_int3, mix_model_int4, mix_model_int5, mix_model_int6)
```

```{r}
#re print the best model and interpret

mix_model_int3<-lmer(DOY~soil_moisture * spp +
                      (1|study_area) + (1|subsite),
                    data = spp_filt)
summary(mix_model_int3)
library(emmeans)
emmeans(mix_model_int3, pairwise ~ spp*soil_moisture)
```


###check assumptions for linear model!!!!!
1. Normality at each X value
2. Homogeneity of variances at each X
3. Independence of observations

```{r}
library(effects)
library(sjPlot)
library(glmmTMB)
plot_model(mix_model_int3, type='diag')
#our model meets all the assumptions
```


### Graphing

```{r}
#simple visualizations of our data

spp_filt %>% 
  ggplot(aes(x=soil_moisture, y=DOY, colour=spp))+
  geom_boxplot()+
  labs(x="Soil Moisture", y="Flowering Day")

spp_filt %>% 
  ggplot(aes(x=spp, y=DOY, colour=soil_moisture))+
  geom_boxplot()+
  labs(x="Species", y="Day of Flowering")+
  theme_classic()#visualization of the t test

avg_doy<-spp_filt %>% 
  group_by(year, soil_moisture) %>% 
  mutate(avg_DOY= mean(DOY))

  ggplot(data = avg_doy, aes(x=year, y=avg_DOY, colour=soil_moisture))+
  geom_point()+
  labs(x="Year", y="Day of Flowering")+
  facet_wrap(~spp)+
    geom_line()+
    theme_classic()
  
  #this shows that theres no change in trend through time^


```
```{r}
#calculating median day of flowering for each species to accompany the boxplot
spp_filt %>% 
  filter(spp == "DRALAC") %>% 
  filter(soil_moisture == "dry") %>% 
  summarize(median = median(DOY))

spp_filt %>% 
  filter(spp == "DRALAC") %>% 
  filter(soil_moisture == "wet") %>% 
  summarize(median = median(DOY))

spp_filt %>% 
  filter(spp == "PAPRAD") %>% 
  filter(soil_moisture == "dry") %>% 
  summarize(median = median(DOY)) 

spp_filt %>% 
  filter(spp == "PAPRAD") %>% 
  filter(soil_moisture == "wet") %>% 
  summarize(median = median(DOY))

spp_filt %>% 
  filter(spp == "PEDHIR") %>% 
  filter(soil_moisture == "dry") %>% 
  summarize(median = median(DOY))

spp_filt %>% 
  filter(spp == "PEDHIR") %>% 
  filter(soil_moisture == "wet") %>% 
  summarize(median = median(DOY))

spp_filt %>% 
  filter(spp == "POLVIV") %>% 
  filter(soil_moisture == "dry") %>% 
  summarize(median = median(DOY))

spp_filt %>% 
  filter(spp == "POLVIV") %>% 
  filter(soil_moisture == "wet") %>% 
  summarize(median = median(DOY))
```



```{r}
#finding flowering time length to compare between wet and dry soils

flower_end<-pheno_data %>% 
  filter(phenophase == "flowerend"| phenophase == "flower") %>% 
  filter(treatment == "CTL") %>% 
  filter(functional_group == "forb") %>% 
  filter(spp == "DRALAC" | spp == "PAPRAD" | spp == "POLVIV" | spp == "PEDHIR") %>% 
  pivot_wider(names_from = phenophase, values_from = DOY) %>% 
  filter(flower != "NA") %>% 
  filter(flowerend != "NA") %>% 
  mutate(flowering_length = flowerend - flower)

  flower_end$soil_moisture <- gsub("moist","wet",flower_end$soil_moisture)
  
#visualizing the data
flower_end %>% 
  ggplot(aes(x=spp, y=flowering_length, colour=soil_moisture))+
  geom_boxplot()+
  labs(x="Species", y="Flowering Length")

#t.tests to compare flowering length in dry and wet soil
dralac<- flower_end %>% 
  filter(spp== "DRALAC")

t.test(flowering_length~soil_moisture, data=dralac)

pedhir<- flower_end %>% 
  filter(spp== "PEDHIR")

t.test(flowering_length~soil_moisture, data=pedhir)

paprad<- flower_end %>% 
  filter(spp== "PAPRAD")

t.test(flowering_length~soil_moisture, data=paprad)

polviv<- flower_end %>% 
  filter(spp== "POLVIV")

t.test(flowering_length~soil_moisture, data=polviv)

```







